---
title: "Breeding-program haplotype workflow (iClass + LGEBV)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Breeding-program haplotype workflow (iClass + LGEBV)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>")
```

# Barley-style end-to-end workflow

This vignette walks through a compact breeding-program haplotype
analysis modeled on the Pandit et al. (2026, *Theoretical and Applied
Genetics*) barley leaf-rust pipeline. It chains three R wrappers
introduced in `rTorchGenomics` 0.4.1:

1. **`tg_iclass()`** — clusters environments by FA-loading polarity
   (Smith et al. 2015 / 2021).
2. **`tg_ld_blocks(method = "r2", tolerance = ...)`** — sliding-window
   haplo-block partition with the new SelectionTools-style `tolerance`
   knob (Wittenburg et al. 2024).
3. **`tg_lgebv()`** — per-block sum of rrBLUP marker effects with a
   sign-based favorable / unfavorable label (Endelman 2011).

```{r}
library(rTorchGenomics)

# One-time setup if you haven't already done it. Idempotent.
tg_install()

bed   <- system.file("extdata", "tiny.bed", package = "rTorchGenomics")
pheno <- system.file("extdata", "tiny_pheno.tsv", package = "rTorchGenomics")
```

## 1. iClass: cluster environments by FA-loading polarity

In a real workflow, you would first fit a multi-environment LMM with
`tg_met_scan(... vg_structure = "fa(k)")` and recover the rotated
factor-analytic loading matrix via the Python side's
`MultiEnvLMM.fa_loadings(null_fit)`. For this vignette we simulate the
output: a 5-environment, 2-factor loading matrix.

```{r}
Lambda <- matrix(
  c( 0.9,  0.4,    # E1 -> "PP"
     0.8,  0.3,    # E2 -> "PP"
     0.7, -0.2,    # E3 -> "PN"
    -0.6,  0.5,    # E4 -> "NP"
    -0.5,  0.4),   # E5 -> "NP"
  nrow = 5, ncol = 2, byrow = TRUE,
  dimnames = list(paste0("E", 1:5), c("F1", "F2"))
)

iclass_result <- tg_iclass(Lambda)
print(iclass_result)
#> IClassResult (n_envs=5, n_factors=2, n_clusters=3, threshold=0.000)
#>   Cluster membership:
#>     PP       -> E1, E2
#>     PN       -> E3
#>     NP       -> E4, E5
```

Environments inside one iClass cluster are expected to be highly
positively genetically correlated; you would now subset your downstream
analyses (BLUE computation, haplo-block scoring) per cluster.

## 2. Haplo-blocks with the new `tolerance` parameter

Run `tg_ld_blocks()` with `method = "r2"` and `tolerance = 2`. With
`tolerance > 0`, up to 2 consecutive weakly linked markers are
permitted inside an otherwise tight block, which often produces
larger, more biologically meaningful blocks than the strict `r2`-greedy
default.

```{r}
blocks_run <- tg_ld_blocks(
  genotype     = bed,
  method       = "r2",
  r2_threshold = 0.7,
  tolerance    = 2,
  output       = tempfile("ldb_")
)
print(blocks_run)
#> LDBlocksRun (method=r2, n_blocks=..., median_size=...kb, runtime=...s)
```

## 3. LGEBV per haplo-block

The LGEBV wrapper consumes the haplo-block partition from step 2 and a
pre-computed BLUE vector (one trait value per genotype). For the
vignette we use the tiny phenotype table directly, but in production
BLUEs would come from an upstream ASReml-R or `tg_met_scan` fit per
iClass cluster.

```{r}
y <- read.table(pheno, header = TRUE, sep = "\t")$y

lgebv_result <- tg_lgebv(
  y                    = y,
  genotype             = bed,
  blocks               = blocks_run,
  h2                   = NULL,         # REML-estimate
  favorable_direction  = "negative",   # resistance trait
  standardize_G        = TRUE
)
print(lgebv_result)
```

## 4. Inspect favorable haplo-blocks

```{r}
to_dataframe(lgebv_result)
#> # A tibble: ... x 8
#>   block_id chrom  start    end n_variants    lgebv block_variance favorable
#>   <chr>    <chr>  <int>  <int>      <int>    <dbl>          <dbl> <lgl>
#> 1 block_0  1        100    300          5    -0.42           0.21 TRUE
#> 2 block_1  1        500    900          6     0.31           0.15 FALSE
#> 3 block_2  2        100    400          4    -0.18           0.08 TRUE
```

The `favorable` column flips when you switch `favorable_direction`:
- `"negative"` (default) — resistance / disease-severity traits where
  lower trait value is better.
- `"positive"` — yield-style traits where higher is better.

## Composing the three wrappers

The full breeding-program workflow is:

```r
# 1. Group environments by FA-loading polarity:
iclass_result <- tg_iclass(Lambda)

# 2. For each iClass cluster, compute BLUEs (upstream) and partition
#    the genome into haplo-blocks:
blocks_run <- tg_ld_blocks(
  genotype = bed, method = "r2",
  r2_threshold = 0.7, tolerance = 2
)

# 3. Score each block within the cluster:
lgebv_result <- tg_lgebv(
  y = blue_vector,
  genotype = bed,
  blocks = blocks_run,
  favorable_direction = "negative"
)
```

## References

- Smith AB *et al.* (2015) *Theoretical and Applied Genetics* 128:55-72.
- Smith AB & Cullis BR (2018) *Euphytica* 214:143.
- Smith AB *et al.* (2021) *Frontiers in Plant Science* 12:737462.
- Endelman JB (2011) *The Plant Genome* 4(3):250-255.
- Bernardo R (2014) *Crop Science* 54(1):68-75.
- Wittenburg D *et al.* (2024) *Theoretical and Applied Genetics* (SelectionTools tolerance).
- Pandit *et al.* (2026) *Theoretical and Applied Genetics* — barley leaf rust LGEBV application.
