---
title: "Single-trait LMM walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-trait LMM walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>")
```

# Single-trait LMM walkthrough

This walkthrough runs an end-to-end GWAS on a tiny PLINK fixture
shipped with the package: 10 samples x 20 variants. The numbers won't
be biologically meaningful; the point is to exercise every step.

## Locating the fixture

```{r}
library(rTorchGenomics)
bed <- system.file("extdata", "tiny.bed", package = "rTorchGenomics")
pheno <- system.file("extdata", "tiny_pheno.tsv", package = "rTorchGenomics")
```

## Pre-flight check

```{r}
v <- tg_validate(genotype = bed, phenotype = pheno)
print(v)
```

Expect `format = bed`, 10 genotype samples, 12 phenotype samples,
10 aligned (the tiny fixture has 2 extra phenotype rows to test
alignment).

## LMM scan

```{r}
r <- tg_lmm_scan(
  genotype = bed,
  phenotype = pheno,
  chunk_size = 10L,       # tiny chunks for the tiny fixture
  maf_min = 0.0,          # don't filter the small variant count
  correction = "bh"
)
print(r)
```

## Inspect results

```{r}
top_hits(r)               # tibble of top variants by p-value
output_files(r)$tsv       # path to the full result table
r@lambda_gc               # lambda_GC (genomic inflation)
```

## Plot

```{r}
mh <- tg_manhattan(r)
qq <- tg_qq(r)

# Save to disk.
ggplot2::ggsave("manhattan.png", mh, width = 8, height = 4)
ggplot2::ggsave("qq.png", qq, width = 6, height = 6)
```

## Where to go from here

- Try `tg_glm_scan()` with `family = "binary"` for case/control studies.
- Use `tg_ld_blocks()` to map haplotype blocks before downstream
  clumping.
- Pipe sumstats into `tg_pgs_fit()` for polygenic scoring.
- For multi-trait or multi-environment designs, see `?tg_mvlmm_scan`,
  `?tg_mtmet_scan`, etc.
