---
title: "Post-GWAS Analysis Toolkit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Post-GWAS Analysis Toolkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, message = FALSE}
library(rTorchGenomics)

# Backend-availability guard (mirrors reticulate-bridge.Rmd): if the
# Python torchgenomics engine isn't importable (CRAN check machines,
# no venv configured), every chunk below is skipped rather than
# erroring. `rTorchGenomics`'s .onLoad() already points reticulate at
# the managed `r-rtorchgenomics` venv when it exists, so a plain
# availability check is enough here.
have_backend <- requireNamespace("reticulate", quietly = TRUE) &&
  reticulate::py_module_available("torchgenomics")

knitr::opts_chunk$set(
  eval = have_backend,
  collapse = TRUE,
  comment = "#>"
)
```

```{r backend-note, echo = FALSE, results = "asis", eval = TRUE}
if (!have_backend) {
  cat(
    "**Python backend not available in this environment** -- every code ",
    "chunk below is shown but was NOT executed. Run `tg_install()` once ",
    "to set up the managed Python runtime, then re-knit.\n",
    sep = ""
  )
}
```

# Post-GWAS Analysis Toolkit

You've run a GWAS and have summary statistics (or a fitted model) in hand.
This vignette tours the post-GWAS facade in `rTorchGenomics` -- the same
`tg_*` functions available from the CLI and from Python's
`torchgenomics.api` -- covering LD clumping, fine-mapping, heritability,
LDSC, power / winner's-curse diagnostics, meta-analysis, multi-ancestry
meta-regression, colocalization, Mendelian randomization, SMR/HEIDI,
gene-set enrichment, and polygenic scoring. Every section below is
executed live against tiny synthetic fixtures (except `tg_annotate_hits`,
which needs live NCBI network access and is shown reference-only).

For the Python-side tour of the same facade with representative output,
see `docs/tutorials/post-gwas-toolkit.md` in the TorchGenomics repository.

## Fixtures

Rather than re-implement the fixture builders in R, this vignette reuses
the verified Python fixture builders from
`examples/python/post_gwas_toolkit.py` via `reticulate::import_from_path()`.
This guarantees the data fed to every `tg_*` call below is byte-identical
to the fixtures used in the (separately verified) Python tutorial.

```{r fixtures}
# Walk up from the current working directory to find the repo's
# examples/python/ directory -- robust to knitr's working directory
# being the vignette's own folder (the usual case) or the repo root.
.find_up <- function(rel, start = getwd(), max_up = 8) {
  d <- normalizePath(start, mustWork = FALSE)
  for (i in 0:max_up) {
    cand <- file.path(d, rel)
    if (file.exists(cand)) return(normalizePath(cand))
    parent <- dirname(d)
    if (identical(parent, d)) break
    d <- parent
  }
  stop(sprintf("Could not locate '%s' by walking up from '%s'.", rel, start))
}

examples_dir <- .find_up(file.path("examples", "python"))
mod <- reticulate::import_from_path("post_gwas_toolkit", path = examples_dir)

tmp_dir <- tempfile("tg_postgwas_vignette_")
dir.create(tmp_dir)
py_path <- reticulate::import("pathlib")$Path(tmp_dir)

fx <- mod$build_clump_ldsc_fixture(py_path)
gwas_clump_ldsc <- as.character(fx[[1]])
geno_clump_ldsc <- as.character(fx[[2]])

gwas_power_wc <- as.character(mod$build_power_wc_fixture(py_path))

fx <- mod$build_finemap_fixture(py_path)
finemap_ss <- as.character(fx[[1]])
finemap_ld <- as.character(fx[[2]])

fx <- mod$build_hess_fixture(py_path)
gwas_hess <- as.character(fx[[1]])
ld_hess <- as.character(fx[[2]])
regions_hess <- as.character(fx[[3]])

fx <- mod$build_coloc_meta_fixture(py_path)
sumstats_a <- as.character(fx[[1]])
sumstats_b <- as.character(fx[[2]])
sumstats_c <- as.character(fx[[3]])

fx <- mod$build_mr_fixture(py_path)
mr_exposure <- as.character(fx[[1]])
mr_outcome <- as.character(fx[[2]])
mr_true_effect <- as.numeric(fx[[3]])

fx <- mod$build_mrmega_fixture(py_path)
mrmega_paths <- vapply(fx, as.character, character(1))

fx <- mod$build_smr_fixture(py_path)
smr_gwas <- as.character(fx[[1]])
smr_eqtl <- as.character(fx[[2]])
smr_gene_map <- as.character(fx[[3]])

fx <- mod$build_enrichment_fixture(py_path)
enrich_gwas <- as.character(fx[[1]])
enrich_genes <- as.character(fx[[2]])
enrich_sets <- as.character(fx[[3]])

fx <- mod$build_pgs_fixture(py_path)
pgs_sumstats <- as.character(fx[[1]])
pgs_ld_ref <- as.character(fx[[2]])
pgs_target_geno <- as.character(fx[[3]])
```

## LD clumping (`tg_clump`)

LD clumping collapses correlated significant hits down to independent
index variants. The fixture has 3 LD blocks of 10 SNPs each, with one
strongly-significant "causal" SNP planted per block.

```{r clump}
r_clump <- tg_clump(
  sumstats = gwas_clump_ldsc, genotype = geno_clump_ldsc,
  output = file.path(tmp_dir, "clump_run"),
  p_threshold = 1e-3, r2 = 0.1, window_kb = 250
)
print(r_clump)
print(r_clump@clumps)
```

## Fine-mapping (`tg_finemap`)

`tg_finemap` runs the SuSiE-RSS engine (the same validated pipeline as
`bayes-scan-rss`) and returns per-variant PIPs plus credible-set summaries.
It expects the uppercase `SNP/CHR/BP/A1/A2/BETA/SE/N` sumstats schema and
a pre-built LD reference carrying matching SNP ids.

```{r finemap}
r_finemap <- tg_finemap(
  sumstats = finemap_ss, ld_ref = finemap_ld,
  max_num_causal = 2L, threads = 1L,
  output = file.path(tmp_dir, "finemap_out.tsv")
)
print(r_finemap)
print(r_finemap@credible_sets)
```

## Local heritability and genetic correlation (`tg_hess`)

HESS estimates local (region-level) SNP heritability from GWAS summary
statistics and an LD matrix; passing a second GWAS (`gwas2 =`) switches
to local genetic correlation instead. The fixture has 4 contiguous
10-SNP regions on chr1, with a heritability signal planted in region 1.

```{r hess-h2}
r_hess_h2 <- tg_hess(
  gwas_hess, ld_hess, regions_hess,
  output = file.path(tmp_dir, "hess_h2.tsv")
)
print(r_hess_h2)
print(r_hess_h2@results)
```

```{r hess-rg}
r_hess_rg <- tg_hess(gwas_hess, ld_hess, regions_hess, gwas2 = gwas_hess)
print(r_hess_rg)
```

## SNP heritability and genetic correlation via LD score regression (`tg_ldsc` / `tg_ldsc_rg`)

`tg_ldsc` and `tg_ldsc_rg` are auto-generated `CliRun` wrappers over the
`ldsc` / `ldsc-rg` CLI subcommands -- they compute LD scores in-process
from a genotype panel and write a small text report to
`<output>.ldsc.txt` / `<output>.ldsc_rg.txt`. **Caveat:** with only 30
SNPs the block-jackknife falls back to a single block, so the printed SE
is `NaN` below -- this demonstrates the call mechanics on toy data, not a
statistically meaningful h2/rg estimate (real runs use thousands of SNPs
across hundreds of jackknife blocks).

```{r ldsc}
ldsc_out <- file.path(tmp_dir, "ldsc_run")
r_ldsc <- tg_ldsc(
  sumstats = gwas_clump_ldsc, genotype = geno_clump_ldsc,
  output = ldsc_out
)
print(r_ldsc)
cat(readLines(paste0(ldsc_out, ".ldsc.txt")), sep = "\n")
```

```{r ldsc-rg}
ldsc_rg_out <- file.path(tmp_dir, "ldsc_rg_run")
r_ldsc_rg <- tg_ldsc_rg(
  sumstats1 = gwas_clump_ldsc, sumstats2 = gwas_clump_ldsc,
  genotype = geno_clump_ldsc, output = ldsc_rg_out
)
print(r_ldsc_rg)
cat(readLines(paste0(ldsc_rg_out, ".ldsc_rg.txt")), sep = "\n")
```

## GWAS detection power (`tg_power`)

Per-variant statistical power, non-centrality parameter, and (with
`power_curve = TRUE`) a min-detectable-effect-size envelope over an
allele-frequency grid.

```{r power}
r_power <- tg_power(
  gwas_power_wc, power_curve = TRUE,
  output = file.path(tmp_dir, "power_out.tsv")
)
print(r_power)
print(r_power@results)
```

## Winner's-curse correction (`tg_winners_curse`)

Significant hits' effect sizes are upward-biased by the selection that
found them ("winner's curse"). `tg_winners_curse` supports three
de-biasing methods; the fixture has two strong hits (subject to the
curse) plus four null/weak SNPs.

```{r winners-curse}
for (method in c("conditional_likelihood", "fiqt", "bootstrap")) {
  r_wc <- if (method == "bootstrap") {
    tg_winners_curse(gwas_power_wc, method = method, n_boot = 200L, seed = 1L)
  } else {
    tg_winners_curse(gwas_power_wc, method = method)
  }
  cat(sprintf("[%s] ", method))
  print(r_wc)
}
```

## Meta-analysis (`tg_meta`)

Fixed-effect meta-analysis of two "studies" over the same 50-SNP region,
sharing one strong causal SNP.

```{r meta}
r_meta <- tg_meta(
  c(sumstats_a, sumstats_b), method = "fixed",
  output = file.path(tmp_dir, "meta_run")
)
print(r_meta)
print(head(r_meta@results))
```

## Multi-ancestry meta-regression (`tg_mr_mega`)

MR-MEGA regresses effect-size heterogeneity onto axes of genetic
variation across populations, rather than assuming a single fixed/random
effect. The fixture has 4 ancestry populations sharing a common causal
effect plus ancestry-specific perturbation.

```{r mrmega}
r_mrmega <- tg_mr_mega(
  mrmega_paths, n_axes = 1L,
  output = file.path(tmp_dir, "mrmega_out.tsv")
)
print(r_mrmega)
print(head(r_mrmega@results))
```

## Colocalization (`tg_coloc`)

Pairwise Giambartolomei-style coloc (PP.H0-H4) for two traits, plus
hyprcoloc for N (here 3) traits sharing a causal SNP.

```{r coloc-pairwise}
r_coloc <- tg_coloc(
  sumstats_a, sumstats2 = sumstats_b, method = "pairwise",
  output = file.path(tmp_dir, "coloc_pairwise.tsv")
)
print(r_coloc)
print(r_coloc@table)
```

```{r coloc-hyprcoloc}
r_hypr <- tg_coloc(c(sumstats_a, sumstats_b, sumstats_c), method = "hyprcoloc")
print(r_hypr)
```

## Two-sample Mendelian randomization (`tg_mr`)

`method = "all"` runs the full sensitivity panel (IVW / Egger /
weighted-median / MR-PRESSO) in one call. The fixture simulates a true
causal effect from exposure to outcome, printed at the end of the chunk
below for comparison against each method's estimate.

```{r mr}
r_mr <- tg_mr(
  mr_exposure, mr_outcome, method = "all",
  n_boot = 200L, n_perm = 200L, seed = 42L,
  output = file.path(tmp_dir, "mr_out.tsv")
)
print(r_mr)
print(r_mr@results)
cat(sprintf("(simulated causal effect = %.2f)\n", mr_true_effect))
```

## SMR + HEIDI (`tg_smr`)

Tests whether a GWAS signal is mediated by gene expression (SMR), then
uses HEIDI to distinguish a single shared causal variant from linkage
between distinct GWAS/eQTL signals.

```{r smr}
r_smr <- tg_smr(
  smr_gwas, smr_eqtl, smr_gene_map,
  output = file.path(tmp_dir, "smr_out.tsv")
)
print(r_smr)
print(r_smr@results)
```

## Gene-set enrichment (`tg_gene_set_enrichment`)

A MAGMA-style competitive gene-set test: SNPs are assigned to genes,
gene-level statistics are aggregated, and gene sets are tested for
enrichment relative to the genomic background. The fixture has 6 genes
(2 strong, 4 null) and one enriched set ("hot" = the 2 strong genes).

```{r enrichment}
r_enrich <- tg_gene_set_enrichment(
  enrich_gwas, enrich_genes, enrich_sets,
  output = file.path(tmp_dir, "enrich_out.tsv")
)
print(r_enrich)
print(r_enrich@results)
```

## Polygenic scores (`tg_pgs_fit` / `tg_pgs_score`)

`tg_pgs_fit` estimates per-SNP PGS weights from summary statistics and
an LD reference panel (here the fast C+T method, suitable for a toy
fixture); `tg_pgs_score` applies fitted weights to a target genotype
panel to produce per-individual scores.

```{r pgs-fit}
pgs_weights <- file.path(tmp_dir, "pgs_weights.tsv")
r_pgs_fit <- tg_pgs_fit(
  pgs_sumstats, pgs_ld_ref, pgs_weights,
  method = "ct", clump_p = 0.5, clump_r2 = 0.9, device = "cpu"
)
print(r_pgs_fit)
```

```{r pgs-score}
r_pgs_score <- tg_pgs_score(
  pgs_target_geno, pgs_weights, file.path(tmp_dir, "pgs_scores.tsv"),
  device = "cpu"
)
print(r_pgs_score)
```

## Gene annotation (`tg_annotate_hits`) -- reference only

`tg_annotate_hits` looks up genes near GWAS hits via the live NCBI
Datasets/E-utilities APIs. It needs network access and is **not
executed** in this vignette; the call shape below is reference only.

```{r annotate, eval = FALSE}
r_annotate <- tg_annotate_hits(
  gwas_power_wc, crop = "maize", p_threshold = 5e-8,
  window_up = 50000L, window_down = 50000L,
  output = file.path(tmp_dir, "annotate_out")
)
print(r_annotate)
```

## Where to go next

Every function above also has a CLI subcommand
(`torchgenomics <subcommand> --help`) and a Python `torchgenomics.api`
equivalent -- see `docs/tutorials/post-gwas-toolkit.md` for the Python
tour with representative output.

```{r cleanup, include = FALSE}
unlink(tmp_dir, recursive = TRUE, force = TRUE)
```
