---
title: "How rTorchGenomics talks to Python"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How rTorchGenomics talks to Python}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>")
```

# How rTorchGenomics talks to Python

rTorchGenomics doesn't reimplement TorchGenomics -- it wraps the Python
engine via [reticulate](https://rstudio.github.io/reticulate/). This
note explains the architecture for advanced users who want to extend
the package or debug the bridge.

## The setup

`tg_install()` creates a Python virtualenv called `r-rtorchgenomics`
and installs `torchgenomics[mcp]` pinned to the matching R-package
version. The venv lives in reticulate's standard location
(`~/.virtualenvs/r-rtorchgenomics/` on Linux/macOS).

On package load, `.onLoad()` tells reticulate to prefer this venv if
it exists. Reticulate doesn't actually import Python until the first
`tg_*()` call -- startup stays fast.

## The first call

The first `tg_*()` call in an R session triggers `tg_py()`, which
calls `reticulate::import("torchgenomics")`. This spawns Python
in-process via the Python C API, loads torch, numpy, and torchgenomics.
Wall-clock cost: ~1-3 seconds, one-time per session.

Subsequent calls re-use the cached module reference -- direct R to Python
function calls with microsecond overhead for arg conversion.

## Per-call flow

```
tg_lmm_scan("data.bed", "pheno.tsv")
    |
    v  R-side arg validation
bridge_call("lmm_scan", list(genotype = "data.bed", ...))
    |
    v  reticulate
py$api$lmm_scan(**args)
    |
    v  Python: actual compute
torchgenomics.api.lmm_scan(...) -> ScanRun dataclass
    |
    v  result.to_dict() in Python -> JSON-safe dict
    |
    v  reticulate::py_to_r() -> R nested list
    |
    v  ScanRun$new_from_dict() -> S4 ScanRun
```

No subprocess. No JSON over stdio. No MCP. Reticulate hosts Python in
the R process; the bridge is a function call.

## Error handling

Python exceptions surface as R conditions with the Python class
hierarchy attached. rTorchGenomics maps them to three R classes:

- `tg_no_python` -- venv not configured.
- `tg_input_error` -- Python ValueError, TypeError, FileNotFoundError.
- `tg_runtime_error` -- anything else.

You can catch them with `tryCatch`:

```{r}
tryCatch(
  tg_lmm_scan("missing.bed", "pheno.tsv"),
  tg_input_error = function(e) {
    message("Bad input: ", conditionMessage(e))
    NULL
  }
)
```

## Going deeper

To inspect or modify the Python module from R:

```{r}
py <- rTorchGenomics:::tg_py()
py$api                # the api facade module
py$models             # low-level models module
py$`__version__`      # torchgenomics version
```

You're calling the same Python objects the LLM-side MCP tools call.
The MCP server (separate Python process for LLM clients) is untouched
by R -- it's a parallel surface, not the bridge.
